---
title: "comaprison"
author: "Yelon"
date: "2026-04-14"
output: html_document
---

```{r}
library(Seurat)
library(ggplot2)
library(patchwork)
library(RColorBrewer)
library(dplyr)
library(pheatmap)
library(ggraph)

set.seed(805181) ## for repeatability of cluster mapping.
```

## 0- load data set

-   preapre only FibHSE and PmtHEE

```{r}
## RUN first "Figure_S7C_S6A_2" to generate "R11_2_kc_fib_mel_in_all_cells.Rdata"
load("R11_2_kc_fib_mel_in_all_cells.Rdata")
# selected_study_cells

DefaultAssay(selected_study_cells)<- "RNA"

# select only FsEpi and PmtHEE
cell_id <- subset(selected_study_cells@meta.data, 
                  !study %in% c("FsEpi")) # 22475 cells

## keep only FibHSE and PmtHEE
selected_obj <- subset(selected_study_cells, cells=row.names(cell_id))
```

## 1- Calculate DEG between KC

-   two conditions: FsEpi vs. PmtHEE.

-   Select only KC cells.

```{r}
## define the only cell KC states
kc_seurat= selected_obj[,selected_obj$Cell_states  %in% c("BAS","SPN","GRN")]

## degine all cells as KC
kc_seurat$Condition <- "KER"

## 1_add new meta.data, cell_type + condition
kc_seurat@meta.data$Condition=paste(kc_seurat@meta.data$study,kc_seurat@meta.data$Condition,sep = "_")

table(kc_seurat$Condition) ## check 
  # FibHSE_KER PmtHEE_KER 
     # 1796       4138 
```

### [save]

```{r}
save(kc_seurat, file = "R16_1_kc_DEG_for_PmtHEE_FibHSE.Rdata")
```

## 2- DEG calculation

```{r}
Idents(kc_seurat)="Condition" # for comparison 1

## 2- for Inactivated vs. Naive. [Comparison 1] [each cell types]
  DEGs_kc <- FindMarkers(
    kc_seurat, 
    logfc.threshold = 0.25,  ## logfc.threshold = 0.25 default
    min.pct = 0.25, #logfc no filter, pct set as low as 0
    only.pos = F, 
    test.use = "wilcox",
    min.cells.feature = 3, # more than 3 cells express genes
 		min.cells.group = 10, ## the cell more than 10 ,default is 3
  	pseudocount.use = 1,
    ident.1="FibHSE_KER",
    ident.2="PmtHEE_KER")

## add the gene column 
  DEGs_kc$gene=rownames(DEGs_kc)
  
```

## 3- Check which condition is higher

```{r}
## test highest genes: FTH1 for FibHSE_KER vs. PmtHEE_KER
custom_colors = c( FibHSE = "#FF7878",
                   PmtHEE= "#3859FF")

p1<-VlnPlot(object = kc_seurat, 
                 features = c("FTH1"), 
                 split.by = "study", 
                 group.by = "Condition",
                 pt.size = 0, 
                 stack = F, 
                 flip = T,
                 cols=custom_colors) 
ggsave("R16_3_DEG_vlnplot_check_FTH1.pdf",
       plot = p1,
       width = 5, 
       height =5)

# So the ident_1 is top number for DEG calcualtion, the positive log means "indet.1" is bigger and higher than ident.2
```

## 4- Annotate the DEG table and prepare the gene list

```{r}
  DEGs_kc$deg_comparison=ifelse(DEGs_kc$avg_log2FC > 0,
                                "FibHSE_KER_up","PmtHEE_KER_up")
      # if log2FC >O. define as FibHSE_up
      # if <0, define as PmtHEE_up"
  
  DEGs_kc_top=DEGs_kc%>%filter(p_val_adj < 0.01) #p_val_adi, significant
  
##> extract top200 genes for each condition
    ## for FibHSE_top200
subset(DEGs_kc_top, deg_comparison %in% c("FibHSE_KER_up"))-> FibHSE_up
FibHSE_up %>%group_by(deg_comparison) %>%top_n(n = 200, wt = avg_log2FC) -> FibHSE_200_up
    ## for PmtHEE_top200
subset(DEGs_kc_top, deg_comparison %in% c("PmtHEE_KER_up"))-> PmtHEE_up
PmtHEE_up %>%group_by(deg_comparison) %>%top_n(n = -200, wt = avg_log2FC) -> PmtHEE_200_up

##> combine them all together as single table
rbind(FibHSE_200_up,
      PmtHEE_200_up)-> combined_matirx

## save it in list() for eahc condition, then use it for BP enrichement.
top200_list <- split(combined_matirx$gene, 
                     combined_matirx$deg_comparison) # how to split the genes, using conditions
```

### [save] top200_gene_list for each condition

```{r}
save(top200_list, file = "R16_4_DEG_top200_list_for_GO_BP.Rdata")
```

## 5- GO_BP enrichment comparison

### 5.1 **Comparing multiple gene lists for BP enrichment**

```{r}
library(org.Hs.eg.db)
library(clusterProfiler)

ck <- compareCluster(geneCluster = top200_list,  # Two list
                     fun = enrichGO,
                     OrgDb = org.Hs.eg.db,
                     keyType="SYMBOL",
                     ont = "BP",
                     pAdjustMethod = "BH",
                     pvalueCutoff = 0.05,
                     qvalueCutoff = 0.05
                     )
```

### [FigS7B] 5.2 visualization

```{r}
p1<-dotplot(ck,
        color = "p.adjust",
        showCategory = 10,
        font.size = 8, # how big is the annotation on axis.
        label_format = 40) ## how long is the y alble lenght

ggsave("FigS7B_DEG_GO_BP_PmtHEE_FibHSE_dotplot.pdf",
       plot = p1,
       width = 6, 
       height =5)
```
