---
title: "figures"
author: "Yelon"
date: "2026-03-06"
output:
  pdf_document: default
  html_document: default
---

-   R 4.4.1

```{r}
library(CellChat)
# install.packages("devtools")
# devtools::install_github("immunogenomics/presto")
# devtools::install_github("jinworks/CellChat")

library(exFINDER)
## devtools::install_github("xyang2uchicago/BioTIP"), as dependencies
# library("devtools")
#library(org.Mm.eg.db)
# BiocManager::install("org.Mm.eg.db"), as dependencies
# BiocManager::install("org.Dr.eg.db")
## devtools::install_github("ChanghanGitHub/exFINDER")
# devtools::install_local("exFinder_data/exFINDER.zip") 

library(Seurat)
library(ggplot2)
library(patchwork)
library(RColorBrewer)
library(dplyr)
library(pheatmap)
library(ggraph)

set.seed(805181) ## for repeatability of cluster mapping.
```

## 1- Seurat input [All studies including KC/Fib/MEL]

```{r}
##> input new all cell object
load("processed_annotated_seurat.Rdata") # new cleaned seurat obj
processed_seurat-> selected_obj  # all cell

##> meta check
table(selected_obj$study)
  # FibHSE  FsEpi PmtHEE 
  #  2248  16388   4139 

table(selected_obj$cell_type_0.7)
  # BAS-I  BAS-II BAS-III  BAS-IV   BAS-V    Endo     Fib   GRN-I  GRN-II GRN-III      HF      LC     MEL 
  # 902     563    1651    2446    2415     114     451    1168    2220    1185      75     111     987 
  #SPN-I  SPN-II SPN-III  SPN-IV 
  # 3026     653    1829    2979 
```

```{r}
# select all KC, Fib and MEL cells
cell_id <- subset(selected_obj@meta.data, 
                  !cell_type_0.7 %in% c("HF", "LC","Endo")) # 22475 cells

## extract the cells along the defined ID 
selected_study_cells <- subset(selected_obj, cells=row.names(cell_id))

# check again the cells
table(selected_study_cells$cell_type_0.7)
    #   BAS-I  BAS-II BAS-III  BAS-IV   BAS-V     Fib   GRN-I  GRN-II GRN-III     MEL   SPN-I  SPN-II SPN-III 
    #902     563    1651    2446    2415     451    1168    2220    1185     987    3026     653    1829 
    # SPN-IV 
    # 2979 
```

## 2- Add extra cell_states

```{r}
## add the KC states for KCs
selected_study_cells-> selected_obj
#> get the cluster number "as character"
cluster_number = names(table(selected_obj$cell_type_0.7))
cluster_number = cluster_number[1:14]
    ##>  [1] "BAS-I"   "BAS-II"  "BAS-III" "BAS-IV"  "BAS-V"   "Fib"     "GRN-I"   "GRN-II"  "GRN-III" "MEL"    
    ## [11] "SPN-I"   "SPN-II"  "SPN-III" "SPN-IV" 

  #> make corresponding cluster annotation.
annotation_cluster = c(
"BAS",
"BAS",
"BAS",
"BAS",
"BAS",
"Fib",
"GRN",
"GRN",
"GRN",
"MEL",
"SPN",
"SPN",
"SPN",
"SPN")

# make the corresponding annotation table with cell_cluster
cluster_anno = data.frame(cluster_number, annotation_cluster, stringsAsFactors = F)

# add empty column for meta.data
selected_obj@meta.data$Cell_states = "NA"

# fill the cell types by cluster number comparison
for(i in 1:nrow(cluster_anno)) { 
selected_obj@meta.data[which(selected_obj@meta.data$cell_type_0.7 == cluster_anno$cluster_number[i]),'Cell_states'] <- cluster_anno$annotation_cluster[i] 
  }

#> check the result and performance
table(selected_obj$Cell_states)
    #  BAS  Fib  GRN  MEL  SPN 
    # 7977  451 4573  987 8487 
```

## [save] selected KC/Fib/MEL in all cells

```{r}
selected_obj-> selected_study_cells
save(selected_study_cells,file = "R11_2_kc_fib_mel_in_all_cells.Rdata")
```

### 2.1- Visualization

```{r}
# BiocManager::install("Rsamtools")
# devtools::install_github("zhanghao-njmu/SCP") # Install SCP under R4.4.1

# options(reticulate.conda_binary = "/Users/yelon/miniconda3/bin/conda")
# SCP::PrepareEnv()
options(reticulate.conda_binary = "/Users/yelon/miniconda3/bin/conda",SCP_env_name = "SCP_env") ## need  specify the conda path first before run SCP
library(SCP)
library(ggplot2)
```

```{r}
#> define the color
library(scales)
library(ggsci)

mycolor<-pal_igv("default", alpha = 0.8)(22)
show_col(mycolor)

mycolor= c(
"BAS-I"="#6BD76BCC",
"BAS-II"="#749B58CC"  ,
"BAS-III"="#CDDEB7CC",
"BAS-IV"="#466983CC" ,
"BAS-V"="#5DB1DDCC",
"SPN-I"= "#924822CC",
"SPN-II"= "#D58F5CCC",
"SPN-III"= "#E4AF69CC",
"SPN-IV"= "#F0E685CC" ,
"GRN-I"= "#AE1F63CC",
"GRN-II"= "#802268CC",
"GRN-III"= "#D595A7CC",
"Fib"= "#CE3D32CC",
"MEL" = "#5050FFCC")

# reorder the cell types
selected_obj$cell_type_0.7 <- factor(x = selected_obj$cell_type_0.7, levels = c(
"BAS-I",
"BAS-II",
"BAS-III",
"BAS-IV",
"BAS-V",
"SPN-I",
"SPN-II",
"SPN-III",
"SPN-IV" ,
"GRN-I",
"GRN-II",
"GRN-III",
"Fib",
"MEL"))

# reorder the study
levels(selected_obj@meta.data$study)

# Reorder the factor levels based on your desired order
selected_obj@meta.data$study <- factor(selected_obj@meta.data$study, 
                                       levels = c("FsEpi", 
                                                  "FibHSE", 
                                                  "PmtHEE"))

p1<-CellDimPlot(selected_obj, 
            palcolor = mycolor, # customize the cell color
            pt.size = 0.7, # define the point size
            group.by = "cell_type_0.7", 
            split.by = "study",
            reduction = "UMAP",
            cells.highlight = TRUE,
            legend.position = "right",
            nrow = 1,
            ncol = 3) +
      theme(plot.title = element_text(size = 20))

ggsave(p1,filename = "R11_2.1_new_umap_plot_study_split.pdf",
       width = 15, height =5)
```

```{r}
# selected_study_cells-> selected_obj
mycolor= c(
"BAS"="#759B59",
"SPN"= "#E3AF6A",
"GRN"= "#D595A7",
"Fib"= "#CE3D32CC",
"MEL" = "#5050FFCC")

# reorder the study
levels(selected_obj@meta.data$study)
selected_obj@meta.data$study <- factor(selected_obj@meta.data$study, 
                                       levels = c("FsEpi", 
                                                  "FibHSE", 
                                                  "PmtHEE"))
# reorder the Cell_states
levels(selected_obj@meta.data$Cell_states)
selected_obj@meta.data$Cell_states <- factor(selected_obj@meta.data$Cell_states, 
                                       levels = c("BAS", 
                                                  "SPN", 
                                                  "GRN",
                                                  "Fib",
                                                  "MEL"))

p1<-CellDimPlot(selected_obj, 
            palcolor = mycolor, # customize the cell color
            pt.size = 0.7, # define the point size
            group.by = "Cell_states", 
            split.by = "study",
            reduction = "UMAP",
            cells.highlight = TRUE,
            legend.position = "right",
            nrow = 1,
            ncol = 3) +
      theme(plot.title = element_text(size = 20))

ggsave(p1,filename = "R11_2.1_new_umap_plot_states_split.pdf",
       width = 15, height =5)
```

## 3- Prepare the split data for ExFinder

```{r}
## split the obj to FsEpi and PmtHEE
split_list <- SplitObject(selected_obj, split.by = "study")

split_list[[1]]-> Fib_obj
split_list$FsEpi-> FsEpi_obj
split_list$PmtHEE-> PmtHEE_obj # keep MEL
```

## [save] split seurat obj for ExFINDER

```{r}
save(Fib_obj,FsEpi_obj,PmtHEE_obj,file = "R11_3_split_obj_for_Exfinder.Rdata")
```

## 4- [FsEpi] DEGs for cell states

```{r}
## based on "Cell_states"-> BAS/FIB/GRN/SPN
load("R11_3_split_obj_for_Exfinder.Rdata")
DefaultAssay(FsEpi_obj) <- "RNA"
Idents(FsEpi_obj)="Cell_states"

markers_celltype <- FindAllMarkers(FsEpi_obj, 
                                   only.pos = TRUE, 
                                   min.pct = 0.25, 
                                   logfc.threshold = 0.25,
                                   test.use = "wilcox")

## extract top10 genes
markers_celltype %>%group_by(cluster) %>%top_n(n = 10, wt = avg_log2FC) -> top10_marker_genes_FsEpi

## save it in list() for eahc cluster,create a list where each element is a vector of genes for each cluster
top10_marker_list_FsEpi <- split(top10_marker_genes_FsEpi$gene, top10_marker_genes_FsEpi$cluster)
```

```{r}
## plot

# scale the data
all.genes <- rownames(FsEpi_obj)
FsEpi_obj <- ScaleData(FsEpi_obj, 
                     features = all.genes)

top10<- top10_marker_genes_FsEpi$gene

p1<-DoHeatmap(FsEpi_obj, 
          features = top10) + 
          NoLegend() +
  scale_fill_gradientn(colors = c("#377EB8", "white", "#E41A1C"))
#> Scale for fill is already present.
#> Adding another scale for fill, which will replace the existing scale.
#> 
#> 
ggsave(p1,filename = "R12_4_top10_heatmap_FsEpi.pdf",
       width = 10, height =10)
```

### [save] top10_marker genes [target genes]

```{r}
save(top10_marker_genes_FsEpi,top10_marker_list_FsEpi, 
     file = "R12_4_top10_marker_FsEpi.Rdata")
```

## 5- Load exFINDER database

```{r}
# load exFINDER database
load("exFinder_data/LR_layer1_human.rda")
load("exFinder_data/RTF_layer2_human.rda")
load("exFinder_data/TFT_layer3_human.rda")

exFINDER.H <- list()
exFINDER.H[[1]] <- LR_layer1_human
exFINDER.H[[2]] <- RTF_layer2_human
exFINDER.H[[3]] <- TFT_layer3_human
```

## 6- [PmtHEE] Prepare the inputs for exFINDER analysis

-   for external signal,

    -   Fib-only

    -   or environment, or unknown cells [not present from all cells]

```{r}
# define the obj
selected_obj <- PmtHEE_obj

# defien the factor
selected_obj$Type <- as.factor(selected_obj$Cell_states) # add "Type", very improtant for later analysis

#> [MEL-only] dataset of Fib-only cells for intersted ligand-expressing cells
cell_id <- subset(selected_obj@meta.data, 
                  Cell_states %in% c("MEL")) # 451 cells
mel_only_cells <- subset(selected_obj, cells=row.names(cell_id))

exp.data <- mel_only_cells@assays$RNA@data
meta.data <- mel_only_cells@meta.data ## need column ""Type" for cell labels

##> [KC-only] dataset
cell_id <- subset(selected_obj@meta.data, 
                  !Cell_states %in% c("MEL")) # 1796 cells
kc_only_cells <- subset(selected_obj, cells=row.names(cell_id))

exp.data2 <- kc_only_cells@assays$RNA@data
meta.data2 <- kc_only_cells@meta.data ## need column ""Type" for cell labels

meta.data2$Type <- droplevels(meta.data2$Type) # if still MEL error

#> [All cells] the complete dataset contains all cells, which used for "external signals" from unknown cells or environment.
exp.Exdata <- selected_obj@assays$RNA@data
meta.Exdata <- selected_obj@meta.data
```

### 6.1- Calculate the threshold for each cell states

-   check the threshold for low and high expressed genes

```{r}
## calculate the percentile values of cell groups for determining the cutoff thresholds for “barely expressed genes” and “highly expressed genes” in later analysis
data <- get_percentile(Exp.Data = exp.data2,
               Meta.Data = meta.data2,
               percentile = c(.5, .75, .90))

table(meta.data2$Type)

p1<-ggplot(data=data, aes(x=Type, 
                      y=Ave.Exp., 
                      color=Prob.)) +
  geom_point(size=3) +
  theme_bw() +
    theme(axis.text.x = element_text(angle = 45, hjust = 0.5, vjust = 0.5))

##
ggsave(p1,filename = "R14_6.1_threshold_check_PmtHEE.pdf",
       width = 5, height =4)

## lowly expressed threshold is lower than 0.5
## highly expressed threshold is higher than 0.9 for each cell group.
## Next, we mark those lowly expressed ligands among all cell groups as potential external signals.
```

## 7- [PmtHEE] **Identify all external signals that come from NON-KC cells targeting KC cells**

-   Could be the ligands from Fib or MEL.

-   ligand--target Gene Regulatory Network (GRN) = ltGRN

-   the external signal-target signaling network = (exSigNet)

-   KC is target cells, need KC marker genes.

```{r}
##> Setup target genes (From KC) and infer ltGRN
##> using KC marker genes
target.gene <- c(top10_marker_list_FsEpi$BAS,
                 top10_marker_list_FsEpi$GRN,
                 top10_marker_list_FsEpi$SPN)

##> infer ltGRN, using KC as target cells
ltGRN <- get_ltGRN(Target = target.gene, DB = exFINDER.H)
  # we get all KC-relatd itGRNs now, as ref.
```

#### [save] KC target genes + ltGRN

```{r}
save(target.gene,ltGRN, file = "R12_7_KC_target_genes_ltGRN.Rdata")
```

### 7.1 [MEL+ Env Ligands, KC as targets]

-   Find all ligands not expressed from KC.
-   target KC as input

```{r}
##> Identify external signals and infer exSigNet
##> [KC-only] used here. We are intersted in the ligands of Non-KC cells, here could be only Fib.
  # should very low in KC, may be highly expressed in the Fib cells or external environment
  # based on ltGRN to find out the ligands very low in KC
  # the threshold to degine the low expressed in KC is < 0.072

meta.data2$Type <- as.factor(meta.data2$Cell_states) # ## very important
      # must be "factor" for all cell types.
Pex <- get_potentialex(Graph = ltGRN,
                Exp.Data = exp.data2,
                Meta.Data = meta.data2,
                cutoff = -0.02, # new lower threshold 
                AG.L = NULL)
## we found 609 ligands not expressed in KC
head(Pex)
#> [1] "PLEKHO2" "INHBE"   "ITGA4"   "ITGA9"   "JAM3"    "JAM2"

```

-   We have KC-ltGRN, and external Ligands, next, check the R/TF and target genes on Target cells t infer exSigNet. define the high threshold....
-   lexternal ligand in KC [low], R/TF/Target shoiuld be high in KC

```{r}
# [KC-only] infer exSigNet (R, TF, T are all highly expressed, L comes from the potential external signals)
##> We need the Lignad low [KC], R/TF/Target high in KC.
##> 
exSigNet <- get_exSigNet(Graph = ltGRN,
                       Ligands = Pex, # the ligands must be potential external signals
                       Exp.Data = exp.data2,
                       Meta.Data = meta.data2,
                       par.cutoff = c(0.3, 0.3), #  two threshold for Receptor and TF.
                       # we select the receptors that are highly expressed in the KC
                       AG.R = c("BAS"),  
                       # we select the receptors that are highly expressed in the KC
                       AG.TF = c('BAS',"SPN","GRN"))

## remove the overlap found.
exSigNet <- fix_OverLap(exSigNet)
# #> [1] "No overlap found!"

# check details
get_NetworkDetail(exSigNet) # show the details of exSigNet
  # 
  # Number of ligands:  120
```

-   Check the ligand expression in all cells, KC should be very low

```{r}
# [All-cells] check the external ligands and their expressions via heatmap
exS <- exSigNet$Graph.Node$Gene[exSigNet$Graph.Node$Role == "Ligand"]

p1<-get_heatmap(Gene = exS,
            Exp.Data = exp.data2,
            Meta.Data = meta.data2,
            cellwidth = 20,
            cellheight = 10,
            fontsize_row = 5,
             cluster_rows = TRUE)

ggsave(p1,filename = "R14_7.1_heatmap_ligand_check_PmtHEE.pdf",
       width = 10, height =20)
```

## 8- [PmtHEE] Identify external signals come from env targeting KC cells.

-   *come from external environment* means very low (or zero) expressions in all measured cells (Fib and KC cells).

### 8.1 [Env Ligands, KC as targets]

-   all cells used here

```{r}
# infer potential external signals from ltGRN: low expression in measured cells [All cells], here predict only ligands
Pex.2 <- get_potentialex(Graph = ltGRN,
                Exp.Data = exp.data2,
                Meta.Data = meta.data2,
                cutoff = -0.02, # lower threshold
                AG.L = NULL) # all from env ligand
head(Pex.2)
#> [1] "PLEKHO2" "INHBE"   "ITGA4"   "ITGA9"   "JAM3"    "JAM2"
```

-   the KC-related exSigNet from env.

```{r}
# [KC-only] infer exSigNet (R, TF, T are all highly expressed, L comes from the potential external signals)
##> We need the Lignad low [KC], R/TF/Target high in KC.
##> 
exSigNet.3 <- get_exSigNet(Graph = ltGRN,
                       Ligands = Pex.2, # the ligands must be potential external signals
                       Exp.Data = exp.data2,
                       Meta.Data = meta.data2,
                       par.cutoff = c(0.3, 0.3), #  two threshold for Receptor and TF.
                       # we select the receptors that are highly expressed in the KC
                       AG.R = c("BAS"),  
                       # we select the receptors that are highly expressed in the KC
                       AG.TF = c('BAS',"SPN","GRN"))


## remove the overlap found.
exSigNet.3 <- fix_OverLap(exSigNet.3)
# #> [1] "No overlap found!"

get_NetworkDetail(exSigNet.3) # show the details of exSigNet
  # Number of ligands:  120 
  # Number of receptors:  32 
```

-   run the following function to select the receptors that only interact with external signals.

```{r}
# any receptors, if inteacted with present cell ligands with be removed.
exSigNet.3 <- filterLR_exSigNet(Graph = exSigNet.3,
                  Exp.Data = exp.data2,
                  DB = exFINDER.H)

get_NetworkDetail(exSigNet.3) # show the details of exSigNet
  # Number of ligands:  24 
  # Number of receptors:  12 
```

```{r}
# check the inferred external signals and their expressions via heatmap
exS.2 <- exSigNet.3$Graph.Node$Gene[exSigNet.3$Graph.Node$Role == "Ligand"]

# visualize on all cells to see ligand expression.
p1<-get_heatmap(Gene = exS.2,
            Exp.Data = exp.data2,
            Meta.Data = meta.data2,
            cellwidth = 20,
            cellheight = 10,
            fontsize_row = 5,
             cluster_rows = TRUE)

ggsave(p1,filename = "R14_8.1_heatmap_env_ligand_PmtHEE.pdf",
       width = 10, height =10)
```

### 8.2 [Env exSigNet, KC as target]

-   no need to selected high expressed ligand, because they are from env, so all should be 0.

```{r}
# calculate the expression levels in Fib, WE ARE intersted only in KC, so still use KC data for target.
exSigNet.4 <- get_NodeValue(Graph = exSigNet.3,
                          Exp.Data = exp.data2, ## target cell KC
                          Meta.Data = meta.data2, ## target cell KC
                          AG.R = c("BAS"),
                          AG.TF = c('BAS',"SPN","GRN"), #
                          AG.T = c('BAS',"SPN","GRN"), # Target genes
                          # if there is no expresison data for ligands, exFINDER will set them to 1.
                          Exp.ExData = NULL, # no data for env
                          Meta.ExData = NULL,
                          AG.ExData = NULL) # no ligand source.

# predict signaling strengths (the default value of Kh is 2)
exSigNet.4 <- get_EdgeWeight(Graph = exSigNet.4, 
               Kh = 2)
```

#### [FigS6A_3] [Visualize exSigNet for env ligand] heatmap

```{r}
# prepare the dataframe for heatmap plot [env ligand-target network]
#> if error, use "get_AveExp(Gene = unique(exSigNet$Graph.Node$Gene),"
exS.2_exp <- get_AveExp(Gene = exSigNet.4$Graph.Node$Gene,
           Exp.Data = exp.data2, # all cells
           Meta.Data = meta.data2)  # all cells

exS.2_exp.HP <- exS.2_exp
exS.2_exp.HP <- exS.2_exp.HP[, -1]
row.names(exS.2_exp.HP) <- exS.2_exp$Gene

# make data.frame includes all cell states
exS.2_exp.HP <- data.frame("BAS" = exS.2_exp.HP$BAS,
                     "SPN" = exS.2_exp.HP$SPN, 
                     'GRN' = exS.2_exp.HP$GRN)

row.names(exS.2_exp.HP) <- exS.2_exp$Gene

# [annotation col] customize the row labels and col labels
annotation_col = data.frame(
  CellType = c("KC", "KC", "KC"))
rownames(annotation_col) = colnames(exS.2_exp.HP)


annotation_row <- data.frame(Role = exSigNet.4$Graph.Node$Role)
row.names(annotation_row) <- row.names(exS.2_exp.HP)
```

-   Plot

```{r}
# heatmap plot
bk <- c(seq(0,1.99,by=0.01), seq(2,4.5,by=0.01))

role_colors <- c(
  "Ligand" = "#FF89FF", 
  "Receptor" = "#FF9288",  
  "TF" = "#01D65C",  
  "Target" = "#82B7FF"   # pink
)

annotation_colors <- list(Role = role_colors,
                          CellType =c("KC"= "#BEC100"))

p <- pheatmap(exS.2_exp.HP, # modify the gene list can modify the heatmap.
         cluster_row = FALSE, 
         cluster_cols = FALSE, 
         cellwidth = 12, 
         cellheight = 12,
         # gaps_row = c(1, 2, 6),
         fontsize_row = 10,
         fontsize_col = 10,
         annotation_row = annotation_row,
         annotation_col = annotation_col,
         annotation_colors = annotation_colors,
         color = colorRampPalette(rev(brewer.pal(n = 7, name ="RdYlBu")))(length(bk)),
         breaks=bk)

ggsave(p,filename = "FigS6A_3_heatmap_exSigNet_for_env_PmtHEE.pdf",
       width = 10, height =30)
```
